---
title: "Perform Multiple Annotations at Once with scBaptism"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Perform Multiple Annotations at Once with scBaptism}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(scBaptism)
```

Vignette Structure

#1: Intro: What does scBaptsim do?

scBaptism is a tool, that enables the user to use multiple existing single cell annotation methods at once (SCINA, CIA, SingleR, Seurat, scPred, scClassify, clustifyr, scmap, CelliD marker based, CelliD reference based), as well as methods to integrate the obtained annotations and options for visualizations. 

Single Cell transcriptomic data can be used to identify different cells inside a sample, the so called annotation. Many different tools are available to automatically perform this process. For researchers this poses the challenge of being familiar with many different tools and how to use them at once. scBaptism offers the possibility to run 10 different annotation methods from 9 existing R packages at once and with the same input data. This makes the process of annotating faster and more accessible. Additionally the resulting annotations can be collapsed into one annotation by majority voting and through a meta cell approach.

2.1: Input format query: What does a user need to provide me? (Installation?)

You can install scBaptism directly from github

```{r setup}
#development version
remotes::install_github("alinajenn/scBaptism@run_anno_tools")
```

The format of your query (aka input data) has to be a SingleCellExperiment object (SCE), containing two assays: 'counts' and 'logcounts'. Dimensionality reductions should also be performed: PCA, tSNE & UMAP. If you have an SCE proceed directly to the next step. If you have a raw expression matrix, you can create a SCE like this: 


```{r setup}

```

2.2 Input format reference and markers_list

Some of the tools included in scBaptism perform the annotation based on a reference and some others use a list of markers. Depending on which tools you want to run one or both if these are needed.

A markers list is required for:

CIA
SCINA
CelliDmk (= marker based Version of CelliD)

#anno_referencebased <- c("SingleR", "Seurat", "clustifyr", "scPred", "scClassify", "scmap", "CelliDref")
  #anno_markerbased <- c("CIA", "SCINA", "CelliDmk")

The markers list must be a nested list with names of genes listed for each cell type.

A reference is required for:

SingleR
Seurat
clustifyr
scPred
scClassify
scmap
CelliDref (= reference based Version of CelliD)

The reference must be a SingleCellExperiment object with an annotation in the colData. You have to pass the name of this column to run_scBaptism in the parameter ref_labs.


```{r setup}
# load example markers list

#for this vignette we will use the labels_main annotation of our example SCE as a reference and perform novel annotation with them on the gene expression data

```

3.1: run_scBaptism

Once you have all the input data ready, running scBaptism can be done by simply calling the main function run_scBaptism()
```{r setup}
#running scBaptism 

result <- run_scBaptism(sce_query = sce_annotated,
                        anno_methods = c("SingleR", "CIA"),
                        reference = sce_annotated,
                        ref_labs = "labels_main",
                        markers_list = markers)

```

-basic -[Seurat Issue note] -passing on parameters -

3.2 Additional parameters

Most of the annotation methods scBaptism can run have many parameters to give the user detailed control over how the annotation is performed. When running run_scBaptism the default parameter setting of these tools are used. If you wish the manually control any specific parameter this is possible and you can simply include them in the function call. 

Some parameters had to be renamed, because otherwise they would get passed on to more then one function. You can still adjust these parameters by using their new name:

Seurat: 
use Seurat_dims instead of dims

CelliD: 
use CelliD_dims instead of dims
use CelliD_features instead of features

scClassify:
use scClassify_features instead of features

clustifyr:
use clustifyr_threshold instead of threshold

scPredict:
use scPredict_threshold instead of threshold

scmap:
use threshold_Cluster instead of threshold (for the cluster based annotation)
use threshold_Cell2Cluster instead of threshold (for the cell based annotation)

```{r setup}
# example: 

```


4: plottings

5: majority voting

6: supercells
