---
title: "Perform Multiple Annotations at Once with scBaptism"
author: Alina Jenn
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Perform Multiple Annotations at Once with scBaptism}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{scBaptism}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Contents

1.  What is scBaptism and what can it do?

2.  Install the package 

3.  Input format query

4.  Input format reference and marker list

5.  Perform the annotations

6.  Configure additional parameters

7.  Plots

8.  Integrate multiple annotations via majority voting

9.  Metacells

10. Session info

## 1 What is scBaptism and what can it do?

scBaptism is a tool, that enables the user to use multiple existing single cell annotation methods at once (SCINA, CIA, SingleR, Seurat, scPred, scClassify, clustifyr, scmap, CelliD marker based, CelliD reference based), as well as methods to integrate the obtained annotations and options for visualizations.

Single Cell transcriptomic data can be used to identify different cells inside a sample, the so called annotation. Many different tools are available to automatically perform this process. For researchers this poses the challenge of being familiar with many different tools and how to use them at once. scBaptism offers the possibility to run 10 different annotation methods from 9 existing R packages at once and with the same input data. This makes the process of annotating faster and more accessible. Additionally the resulting annotations can be collapsed into one annotation by majority voting and through a meta cell approach.

## 2 Install the package

You can install scBaptism directly from github

```{r installation, eval=FALSE}
#development version
remotes::install_github("alinajenn/scBaptism@run_anno_tools")
```

Load the package

```{r setup, eval=TRUE}
library("scBaptism")
```

## 3 Input format query

The format of your query (aka input data) has to be a SingleCellExperiment object (SCE), containing two assays: 'counts' and 'logcounts'. Dimensionality reductions should also be performed: PCA and tSNE for plotting later on. (The tool clustifyr also requires cluster information to be present).

```{r load data}
#we are going to use an SCE from the iUSEiSEE package as an example for this vignette

library(iUSEiSEE)
sce_example <- readRDS(file = system.file("datasets", 
                                          "sce_pbmc3k.RDS", 
                                          package = "iUSEiSEE"))

```

## 4 Input format reference and markers list

Some of the tools included in scBaptism perform the annotation based on a reference and some others use a list of markers. Depending on which tools you want to run one or both if these are needed.

A markers list is required for:

- CIA

- SCINA

- CelliDmk (= marker based Version of CelliD)

The markers list must be a nested list with names of genes listed for each cell type.

A reference is required for:

- SingleR

- Seurat

- clustifyr

- scPred

- scClassify

- scmap

- CelliDref (= reference based Version of CelliD)

The reference must be a SingleCellExperiment object with an annotation in the colData. You have to pass the name of this column to run_scBaptism in the parameter ref_labs.

```{r load reference and markers}

#for this vignette we will use the labels_main annotation of our sce_example as a reference and perform new annotation with them on the gene expression data


#quick look at the reference annotation in the colData, called "labels_main
table(sce_example$labels_main)


# from the reference we can create a markers list using the Seurat package

library(Seurat)
library(dplyr)

myseu <- Seurat::as.Seurat(sce_example)
myseu <- Seurat::ScaleData(myseu)
Seurat::Idents(myseu) <- "labels_main"
seu_all_markers <- Seurat::FindAllMarkers(myseu, 
                                          test.use = "wilcox", 
                                          only.pos = TRUE,
                                          min.pct = 0.25, 
                                          logfc.threshold = 0.25)

top_k_markers <- 50

markers_lists <- seu_all_markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::top_n(n = top_k_markers, wt = avg_log2FC)

markers <- split(markers_lists$gene, markers_lists$cluster)


```

## 5 Perform the annotations

Once you have all the input data ready, running scBaptism can be done by simply calling the main function run_scBaptism():

```{r run annotation}
#running scBaptism 

library(scBaptism)

result <- run_scBaptism(sce_query = sce_example,
                        anno_methods = c("SingleR", "CIA"),
                        reference = sce_example,
                        ref_labs = "labels_main",
                        markers_list = markers)

# if you want to run all available annotation, provide this vector:
anno_methods = c("SingleR", "Seurat", "clustifyr", "scPred", "scClassify", "scmap", "CelliDref", "CIA", "SCINA", "CelliDmk")

```

The return object is your SCE query with the annotation columns added to the colData. All new annotation columns are named following this scheme: "scb_toolname_labels". "scb" stand for scBaptism, so you can easily identify which columns are a result of running run_scBaptism.

[Important Note: Currently all packages that use the Seurat function ScaleData() (Seurat, scPred, CelliDref) take a very long time to run (30+ mins). This is an internal issue with the Seurat package. We did make the developers aware of this issue and asked them to fix it, but it could take some time until it happens. So maybe skip these tools, if you don´t want to wait forever]

## 6 Configure additional parameters

Most of the annotation methods scBaptism can run have many parameters to give the user detailed control over how the annotation is performed. When running run_scBaptism the default parameter setting of these tools are used. If you wish the manually control any specific parameter this is possible and you can simply include them in the function call.

Some parameters had to be renamed, because otherwise they would get passed on to more then one function. You can still adjust these parameters by using their new name:

- Seurat: use Seurat_dims instead of dims

- CelliD: use CelliD_dims instead of dims use CelliD_features instead of features

- scClassify: use scClassify_features instead of features

- clustifyr: use clustifyr_threshold instead of threshold

- scPredict: use scPredict_threshold instead of threshold

- scmap: use threshold_Cluster instead of threshold (for the cluster based annotation), use threshold_Cell2Cluster instead of threshold (for the cell based annotation)

```{r parameters adjustment}
# example: Same call from above with the parameter unassigned_label of SingleR adjusted to a custom name

result_custom_para <- run_scBaptism(sce_query = sce_example,
                        anno_methods = c("SingleR", "CIA"),
                        reference = sce_example,
                        ref_labs = "labels_main",
                        markers_list = markers,
                        unassigned_label = "Unassigned Cell")

```

## 7 Plots

The scBaptism package comes with three plotting function to give you a quick way to look at your data:

- plot_multiple_tSNE() plots the tSNE for each annotation in individual tiles

- plot_alluvial() compares the differences between two annotations in an alluvial plot

- plot_confusion_matirx() compares the differences between two annotations in a confusion matrix

```{r plotting}
# Let's plot our example annotations from SingleR and CIA

tSNE_plot <- plot_multiple_tSNE(sce_query = result,
                               labels_vector = c("scb_SingleR_labels", "scb_CIA_labels"))

tSNE_plot

matrix_plot <- plot_confusion_matrix(sce_query = result,
                                     first_tool = "scb_SingleR_labels",
                                     second_tool = "scb_CIA_labels")

matrix_plot

alluvial_plot <- plot_alluvial(sce_query = result,
                               first_tool = "scb_SingleR_labels",
                               second_tool = "scb_CIA_labels",
                               threshold = 0.02)

alluvial_plot

```

## 8 Integrate multiple annotations via majority voting

If you have multiple annotations and want to have an annotation that represents a consensus of the majority of the tool you can run the function majority_voting(). For each cell it will return the annotation that occurred the most among all of the tools. In case of a tie you can choose one of these tie breaker rules for the parameter tie_breaker:

- 'first': the annotation from the first tool is assigned to this cell
- 'concat': all annotation from all tools get concatenated into one string (separated by "\|")
- 'LCA': annotation assigned is the last common ancestor of all the annotations (via ontoProc package)
- 'na': tie is marked as NA

The input for this function is a SCE that contains multiple annotations. The parameter anno_columns decides which columns the majority voting will be performed on. You can either provide a vector with the name of all annotation columns or you set it to NULL (also the default) and the function will automatically include all annotations obtained with the run_scBaptism function (every column that starts with "scb" to be precise).

The majority_voting function returns the input SCE object with an additional colData column called "scb_majority_labels".

```{r majority vote}
# perform a majority vote on all annotations obtained through scBaptism, with choosing the first element in case of a tie
result_mv <- majority_vote(sce_query = result, 
                               anno_columns = NULL, 
                               tie_breaker = "first")

# plot the majority vote annotation tSNE
scater::plotTSNE(result_mv, color_by = "scb_majority_labels")

```

## 9 Metacells

Another way to gain different insights into your data is to agglomerate your single cell data into so-called metacells. 

Metacells typically capture the expression data of around 10 to 50 single cells. This is a way to reduce noise in your data and also work with a smaller data set. In the scBaptism package, you can turn your single cell level data into metacells data by simply calling the calculate_metacells() function. This returns a SingleCellExperiment object for you with a gene expression matrix of log counts that represent Meta-cells. Additionally, PCA, T, S, and E, and clusters are also calculated for further usage. The function takes two parameters: The first is simply your query data in the form of a SingleCellExperiment object. The second parameter gamma decides the graining level. A graining level of 10 would mean that 10 single cells get combined into one metacell. 


```{r metacells}
sce_metacells <- calculate_metacells(sce_query = sce_example,
                                     gamma = 10)

sce_metacells
```

The function returns to you a SingleCellExperiment object with log counts. These log counts were calculated by averaging the expression data of each single cell contained in the meta-cell. Additionally, PCA, tSNE and clusters were calculated.

## 10 Session Info

```{r sessioinfo}
#| class-output: smaller

sessionInfo()
```
