---
title: "PBMC3k workflow to test the whole package"
output: html_document
date: "2026-06-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Load all neccessary libraries
```{r libraries}
library(scBaptism)
library(iUSEiSEE)
library(Seurat)
library(dplyr)
library(cowplot)

```


## Load PBMC3k dataset in SCE format with main preprocessing steps already done (log-normalization, clustering, t-SNE, PCA)
```{r load SCE}
sce_query <- readRDS(file = system.file("datasets", 
                                        "sce_pbmc3k.RDS", 
                                        package = "iUSEiSEE"))

```

## The labels_main annotation from the example SCE will serve as a reference

From this reference the list of marker genes is build with the Seurat package


```{r }

#reference
ref <- sce_query

#markers list
myseu <- Seurat::as.Seurat(sce_query)
myseu <- Seurat::ScaleData(myseu)
Seurat::Idents(myseu) <- "labels_main"
seu_all_markers <- Seurat::FindAllMarkers(myseu, 
                                          test.use = "wilcox", 
                                          only.pos = TRUE,
                                          min.pct = 0.25, 
                                          logfc.threshold = 0.25)

top_k_markers <- 50

markers_lists <- seu_all_markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::top_n(n = top_k_markers, wt = avg_log2FC)

markers <- split(markers_lists$gene, markers_lists$cluster)


```

## Annotate the data set with all methods included in scBaptism

```{r}
anno_result <- run_scBaptism(sce_query = sce_query,
                            anno_methods = c("SingleR", 
                                             "Seurat", 
                                             "clustifyr", 
                                             "scPred", 
                                             "scClassify", 
                                             "scmap", 
                                             "CelliDref", 
                                             "CIA", 
                                             "SCINA", 
                                             "CelliDmk"),
                             reference = ref,
                             ref_labs = "labels_main",
                             markers_list = markers,
                             clusters = "Cluster")


```


## plotting with integrated visualizations

Plotting of heatmap and alluvial is only done against SingleR for all tools, 
to not end up with an excessive amount of tools

```{r}

#plot all annotations as t-SNE plot, plus the reference annotation
tSNE_plot <- plot_multiple_tSNE(sce_query = anno_result,
                               labels_vector = c("scb_SingleR_labels", 
                                                 "scb_CIA_labels", 
                                                 "scb_Seurat_labels",
                                                 "scb_clustifyr_labels",
                                                 "scb_scPred_labels",
                                                 "scb_scClassify_labels",
                                                 "scb_scmap_labels",
                                                 "scb_SCINA_labels",
                                                 "scb_CelliDref_labels",
                                                 "scb_CelliDmk_labels",
                                                 "labels_main"))

tSNE_plot


#vector with all annotations, except SingleR, so all can be plotted against it
labels_vector = c(
                     "scb_CIA_labels", 
                     "scb_Seurat_labels",
                     "scb_clustifyr_labels",
                     "scb_scPred_labels",
                     "scb_scClassify_labels",
                     "scb_scmap_labels",
                     "scb_SCINA_labels",
                     "scb_CelliDref_labels",
                     "scb_CelliDmk_labels")

# plot heatmaps exemplary for SingleR against all other cell annotations

heatmap_list <- lapply(labels_vector, function(arg){plot_heatmap(sce_query = anno_result,first_tool = "scb_SingleR_labels", second_tool = arg)})


heatmap_plot2 <- cowplot::plot_grid(plotlist = heatmap_list, nrow = 4, ncol = 2)



# plot alluvial plots exemplary for SingleR against all other cell annotations


alluvial_list <- lapply(labels_vector, function(arg){plot_alluvial(sce_query = anno_result,first_tool = "scb_SingleR_labels", second_tool = arg, threshold = 0.02)})

#display all plots in one graphic
alluvial_plot <- cowplot::plot_grid(plotlist = alluvial_list, nrow = 4, ncol = 2)


#demonstrate filtering, at 2% and 10 %
thresholds <- c(0.02, 0.1)

alluvial_list_filter <- lapply(thresholds, function(arg){plot_alluvial(sce_query = anno_result,first_tool = "scb_SingleR_labels", second_tool = "scb_scPred_labels", threshold = arg)})

alluvial_plot_filter <- cowplot::plot_grid(plotlist = alluvial_list_filter, ncol = 2)
```


#perform majority voting
#with all tie breakers and plot against the reference annotation

```{r}
#perform majority voting with all tie breaker options

result_all <- anno_result

result_first <- majority_vote(sce_query = anno_result, 
                           anno_columns = NULL, 
                           tie_breaker = "first")

#rename column to avoid confusion later on
names(SummarizedExperiment::colData(result_first))[which(names(SummarizedExperiment::colData(result_first))=="scb_majority_labels")]="mv_first"

result_LCA <- majority_vote(sce_query = anno_result, 
                           anno_columns = c("scb_SingleR_labels",
                                            "scb_Seurat_labels",
                                            "scb_clustifyr_labels", 
                                            "scb_scPred_labels",
                                            "scb_scClassify_labels",
                                            "scb_scmap_labels",
                                            "scb_CelliDref_labels", 
                                            "scb_CIA_labels",
                                            "scb_SCINA_labels",
                                            "scb_CelliDmk_labels"), 
                           tie_breaker = "LCA")

names(SummarizedExperiment::colData(result_LCA))[which(names(SummarizedExperiment::colData(result_LCA))=="scb_majority_labels")]="mv_LCA"

result_concat <- majority_vote(sce_query = anno_result, 
                           anno_columns = NULL, 
                           tie_breaker = "concat")

names(SummarizedExperiment::colData(result_concat))[which(names(SummarizedExperiment::colData(result_concat))=="scb_majority_labels")]="mv_concat"

result_na <- majority_vote(sce_query = anno_result, 
                           anno_columns = NULL, 
                           tie_breaker = "na")

names(SummarizedExperiment::colData(result_na))[which(names(SummarizedExperiment::colData(result_na))=="scb_majority_labels")]="mv_na"



SummarizedExperiment::colData(result_all)$mv_first <- result_first$mv_first
SummarizedExperiment::colData(result_all)$mv_LCA <- result_LCA$mv_LCA
SummarizedExperiment::colData(result_all)$mv_concat <- result_concat$mv_concat
SummarizedExperiment::colData(result_all)$mv_na <- result_na$mv_na

```


##Metacells

#first: convert the example SCE into Metacells and then annotate it again

#second: Take existing annotation and take it along with the Metacell

```{r}
# convert SCE into metacells, then annotate it again
#graining levels tested: 10 , 20 and 30

metacells_ten <- calculate_metacells(sce_query = anno_result,
                                     annotation_col = NULL,
                                     gamma = 10,
                                     n.pc = 10
                                     )

metacells_twenty <- calculate_metacells(sce_query = anno_result,
                                     annotation_col = NULL,
                                     gamma = 20,
                                     n.pc = 10
                                     )

metacells_thirty <- calculate_metacells(sce_query = anno_result,
                                     annotation_col = NULL,
                                     gamma = 30,
                                     n.pc = 10
                                     )


#annotate the objects with SingleR and CIA as an example

metacells_ten <- run_scBaptism(sce_query = metacells_ten, 
                               anno_methods = c("SingleR", "CIA"), 
                               reference = ref,
                               ref_labs = "labels_main",
                               markers_list = markers)

metacells_twenty <- run_scBaptism(sce_query = metacells_twenty, 
                               anno_methods = c("SingleR", "CIA"), 
                               reference = ref, 
                               ref_labs = "labels_main",
                               markers_list = markers)

metacells_thirty <- run_scBaptism(sce_query = metacells_thirty, 
                               anno_methods = c("SingleR", "CIA"), 
                               reference = ref,
                               ref_labs = "labels_main",
                               markers_list = markers)

```

```{r metecells with annotation}
#now lets try to make metacells and take along the existing annotation from our anno_results
metacells_ten_anno <- calculate_metacells(sce_query = anno_result,
                                     annotation_col = "scb_SingleR_labels",
                                     gamma = 10,
                                     n.pc = 10
                                     )

metacells_twenty_anno <- calculate_metacells(sce_query = anno_result,
                                     annotation_col = "scb_SingleR_labels",
                                     gamma = 20,
                                     n.pc = 10
                                     )

metacells_thirty_anno <- calculate_metacells(sce_query = anno_result,
                                     annotation_col = "scb_SingleR_labels",
                                     gamma = 30,
                                     n.pc = 10
                                     )

```

## Plotting Majority voting and metacells
```{r}
#metacells: graining levels with annotation transferred

sce_vector <- c(metacells_ten_anno, metacells_twenty_anno, metacells_thirty_anno)

MCafter_list <- lapply(sce_vector, function(arg){scater::plotTSNE(object = arg, color_by = "scb_SingleR_labels")})

#display all plots in one graphic
MCafter_plot <- cowplot::plot_grid(plotlist = MCafter_list, ncol = 3, labels = c("Graining level 10", "Graining level 30", "Graining level 30"))
```

